# Global tests

## One-sample problem


```{r fig-slopes_mean}
#| fig-subcap:
#|  - Mean annual slope of the maximum temperature distributions between 1987 and 2016.
#|  - Centered log-ratio transform of the mean annual slope of maximum temperature distributions between 1987 and 2016.
#| fig-height: 3
#| layout-ncol: 2
vnt_mean <- mean(c(vnt$t_max))

plot(vnt_mean) +
  geom_hline(yintercept = 1 / diff(vnt$t_max[[1]]$basis$rangeval), linetype = "dotted") +
  scale_y_continuous(n.breaks = 4) +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )

plot(as.fd(as.list(vnt_mean))) +
  scale_y_continuous(n.breaks = 4) +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "clr"
  )
```

Expliquer davantage l'implémentation (approximation chi-2 etc.) Zhang a
partir de page 107.

Inlcure graphique de la moyenne globale

Here $G=1$ therefore we eliminate the index $g$ from previous notations.
We are given a reference density $\pi_0.$ Of particular interest is the
case $\pi_0$ is the null element of the Bayes space.

### Test hypotheses

$H_0: {\mathbb E}^B (\pi_i) = \pi_0, \forall i=1, \ldots n$. Because
$${\mathbb E}^B (\pi_i) = \pi_0 \Leftrightarrow \operatorname{clr} {\mathbb E}^B (\pi_i) =  \operatorname{clr} \pi_0 \Leftrightarrow {\mathbb E} (\operatorname{clr} \pi_i) =  \operatorname{clr} \pi_0$$
we may use one-sample test in $L^2$ space. Answers the question: "is the
mean density equal to the reference $\pi_0$ ?"

Local test for fixed $x$: $H_0: {\mathbb E}^B (\pi_i)(x) = \pi_0 (x)$?

Two problems:

-   $H_0: \mathbb{E}^B(\pi_1)(x) = \mathbb{E}^B(\pi_2)(x)$ for a given
    point $x$ is not equivalent to
    $\operatorname{clr}\mathbb{E}^B(\pi_1)(x) = \operatorname{clr}\mathbb{E}^B(\pi_2)(x)$
    because the clr involves all x's therefore the Student statistic is
    not appropriate.

-   Meaning of pointwise hypotheses for a density function which is
    defined almost everywhere ?? only makes sense on subintervals, we
    would like to know on which ranges of $x$ the densities $\pi_i$
    differ from $\pi_0$.

### Test statistics

Two options:

-   Norm approach
    $$T_\text{norm}= n \| \operatorname{clr} (\bar{\pi}_{..}) - \operatorname{clr} (\pi_0) \|^2_{L^2_0(a,b)}$$
    Reference: [@kokoszka_introduction_2017]

-   PC approach (Martin, report 7, page 13 et Kokotzka page 269)
    Hotelling's T-square statistic based on functional PCA $T_{PC}.$
    Given a truncation with $p$ terms
    $$T_{PC} = n\sum_{k=1}^p \frac{\langle \operatorname{clr} (\bar{\pi}_{..}) - \operatorname{clr} (\pi_0), v_k \rangle_{L^2_0(a,b)}^2}{\hat \lambda_k,}$$
    where $\lambda_k$ are the eigenvalues and $v_k$ the eigenfunctions
    of FPCA.

Comparison: Kokoszka recommends that the PC approach be used when a very
small number of PCs explain a larger proportion of variability, and to
use the norm approach otherwise.

### Application: changes in temperature distribution over the period 1987-2016

Changes in temperature distributions will be measured through linear
slopes by province.

We perform the one-sample test for the slope sample of $\beta_{gi},$ and
$\pi_0$ equal to the neutral element (uniform density on $(a,b).$ it
allows to answer the question: "are the mean slope densities in the
provinces equal to the Bayes neutral element $\pi_0$" ?

```{r tbl-one_sample}
#| tbl-cap: Test statistics and $p$-values for the global one-sample problem.

one_sample <- function(ddobj, p = 3) {
  n <- ncol(ddobj$coefs)
  d <- nrow(ddobj$coefs)

  pca_fit <- pca.fd(ddobj, nharm = d)
  lambda <- pca_fit$values
  scores <- pca_fit$scores
  v <- pca_fit$harmonics

  # PC test
  tpc <- n * sum(inprod(v[1:p], mean(ddobj))^2 / lambda[1:p])
  tpc_pval <- pchisq(tpc, p, lower.tail = FALSE)

  # Norm test
  tnorm <- n * sum(inprod(v[lambda[1:d] > 0], mean(ddobj))^2)
  tnorm_pval <- imhof(tnorm, lambda[lambda[1:d] > 0])[[1]]

  one_sample_tbl <- tibble(
    Name = c("PC", "Norm"),
    Statistic = c(tpc, tnorm),
    `p-value` = c(tpc_pval, tnorm_pval)
  )
}

kable_format(one_sample(c(vnt$t_max)))
```


## Distributional ANOVA


```{r fig-slopes_regional_means}
#| fig-cap: Regional mean slope and overall slope.
vntr <- vnt |>
  select(!province) |>
  group_by(region, region_name) |>
  summarise(t_max = mean(t_max))

vntr |> plot_funs(t_max, color = region_name) +
  geom_hline(yintercept = 1 / diff(vnt$t_max[[1]]$basis$rangeval), linetype = "dotted") +
  facet_grid(vars(region), axes = "all") +
  scale_color_viridis_d() +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )
```

```{r fig-slopes_regional_means_centered}
#| fig-cap: Regional mean slope, centered by the global mean.
#| fig-height: 3
vntr <- vnt |>
  select(!province) |>
  mutate(t_max = center(t_max)) |>
  group_by(region, region_name) |>
  summarise(t_max = mean(t_max), .groups = "drop") |>
  mutate(t_max = as.fd(t_max))

vntr |> plot_funs(t_max, color = region_name) +
  scale_color_viridis_d() +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Centered log-ratio transform"
  )
```

### Test hypotheses

$H_0: \forall g \quad {\mathbb E}^B (\pi_{gi}) = {\mathbb E}^B (\pi_{1i})$
equivalent to
$H_0: \forall g \quad {\mathbb E} (\text{clr }\pi_{gi}) = {\mathbb E} (\text{clr }\pi_{1i})$

[Chris: with the model definition of
[\[eq:danova_model2\]](#eq:danova_model2){reference-type="eqref"
reference="eq:danova_model2"}, $H_0$ becomes the equality of all group
main-effect to zero
$H_0: \forall g \quad {\mathbb E}^B (\alpha_{g}) = 0_{{\cal B}^2}$
]{style="color: blue"}

### Test statistics

Reference the FANOVA version is [@gorecki_fdanova_2019].

In the forthcoming formulas, to adapt FANOVA to DANOVA, the pointwise
variations $SSB(x)$ and $SSW(x)$ from
[\[eq:SSBloc\]](#eq:SSBloc){reference-type="eqref"
reference="eq:SSBloc"} and
[\[eq:SSWloc\]](#eq:SSWloc){reference-type="eqref"
reference="eq:SSWloc"} will be applied to the clr transforms
$f_{gi}= \operatorname{clr}(\pi_{gi}).$

1.  The $L^2$-norm based test from [@zhang_statistical_2007]. The test
    statistic is given by $$\label{eq:zhang}
        L^2B= \int_a^b SSB(x)dx$$ Note that an alternative formula for
    $L^2B$ when applied to clr of densities is given by
    $$\label{eq:zhang2}
        L^2B= \sum_{g=1}^G n_g \parallel \bar\pi_{g.} - \bar\pi_{..}\parallel^2_{B^2},$$
    which shows that this statistic can be expressed as a norm and is
    therefore invariant under almost-everywhere equality. Moreover this
    norm coincides with the norm in Bayes space of the difference
    between group means and overall mean. The `L2B` implementation uses
    an asymptotic distribution of the test statistic.

2.  The `CS` statistic from [@cuevas_anova_2004] uses pairwise
    differences. When applied to clr of densities, due to the linearity
    of $\operatorname{clr}$, we get $$\label{eq:cuevas}
        CS = \sum_{g<g'} n_g \int_a^b \left( \operatorname{clr} {\bar\pi_{g.}}(x)- \operatorname{clr} {\bar\pi_{g'.}}(x) \right)^2dx =  \sum_{g<g'} n_g  \parallel \bar\pi_{g.} - \bar\pi_{g'.}\parallel^2_{B^2}$$

    The `CS` implementation in the `fdANOVA` package uses a
    heteroscedastic assumption and parametric bootstrap. Note that an
    alternative formula for $CS$ is the sum of the Bayes norm of the
    pairwise differences between groups.

3.  $F$-type tests using both within and between variation, see
    [@zhang_statistical_2011] and [@shen_f_2004]. We choose the `FB`
    implementation (biased-reduced) for the quantile computation under
    the null. $$\label{eq:Fvdb}
        FB = \frac{{\int_a^b SSB(x)dx}/(G-1)}{\int_a^b SSW(x)dx /(n-G)} = \frac{\sum_{g=1}^G n_g \parallel \bar\pi_{g.} - \bar\pi_{..}\parallel^2_{B^2}/(G-1)}{\sum_{g=1}^G \sum_{i=1}^{n_g}  \parallel \pi_{gi} - \bar\pi_{g.}\parallel^2_{B^2}/(n-G)},$$
    Under the null hypothesis, this statistic is asymptotically equal to
    a linear combination of independent $\chi^2$ for the numerator and
    the denominator, which can be approximated by a Fischer distribution
    [@zhang_statistical_2011 Theorem 1 and equation 2.16]. This approach
    coincides with the approach of [@van_den_boogaart_bayes_2014] by the
    second equality in [\[eq:Fvdb\]](#eq:Fvdb){reference-type="eqref"
    reference="eq:Fvdb"}. Note that [@van_den_boogaart_bayes_2014] use a
    bootstrap procedure rather than the above approximation.

4.  `GPF` from [@zhang_one-way_2014]: instead of integrating separately
    the pointwise within variation (resp: between), the test statistic
    integrates the pointwise F-ratio: $$\label{eq:GPF}
        GPF = \int_a^b \frac{SSB(x)/(G-1)}{SSW(x)/(n-G)}dx$$ Under the
    null hypothesis, this statistic is asymptotically equal to a linear
    combination of independent $\chi^2$, which can be approximated by a
    $\chi^2$-distribution [@zhang_one-way_2014 Proposition 1]. In the
    implementation of the `fdANOVA` package, `GPF` is divided by $b-a$,
    and the null distribution is modified accordingly.

5.  `Fmax` from [@zhang_new_2019]: instead of integrating the pointwise
    F-ratio as in `GPF`, they compute its supremum. We consider the
    implementation `Fmaxb` which bootstraps the distribution under the
    null hypothesis. $$\label{eq:Fmax}
        F_{\max} = \sup_{x \in (a,b)} \frac{SSB(x)/(G-1)}{SSW(x)/(n-G)}$$
    The statistic $F_{\max}$ is the only one that is not invariant under
    almost-everywhere equality: changing one value of a density in the
    dataset might change the value of $F_{\max}$. Note also that the
    statistics $GPF$ and $F_{\max}$ cannot be written straightforwardly
    in terms of norms in the Bayes space.

Argument: in [@gorecki_fdanova_2018], they claim that the GPF test is
more powerful than the F-type tests (page 5).

### Application: regional changes in temperature distribution over the period 1987-2016

$\pi_{git}$ maximum daily temperature density f for year $t$. We first
apply for each province the slope estimation and obtain the set of slope
densities $\beta_{gi=}.$ Then we test whether these slope densities vary
across regions. Table [1](#tab:my_label){reference-type="ref"
reference="tab:my_label"} summarizes the results for the above tests
which all conclude that there is a difference in the way the regionial
temperature densities evolve in time.

```{r tbl-slopes_danova_global}
#| tbl-cap: Test statistics and $p$-values for DANOVA.
#| warning: false

vnte <- vnt |>
  mutate(t_max = as.fd(t_max)) |>
  eval_funs(t_max, n = 1401) |>
  select(province, x, y) |>
  pivot_wider(names_from = province, values_from = y) |>
  column_to_rownames("x")

vnt_province_region <- vietnam_provinces$region
names(vnt_province_region) <- vietnam_provinces$province

stat_list <- list("GPF", "Fmaxb", "CS", "L2B", "FB")

res <- fanova.tests(vnte, vnt_province_region[colnames(vnte)],
  test = stat_list,
  params = list(
    paramCS = 100,
    paramFmaxb = 100,
    paramTRP = list(B.TRP = 100)
  )
)

danova_tbl <- res[as_vector(stat_list)] |>
  sapply(function(x) c(`Statistic` = x$stat, `p-value` = x$pval)) |>
  t() |>
  as.data.frame() |>
  rownames_to_column("Name")

kable_format(danova_tbl)
```

## One-sample problem for each group

```{r tbl-one_sample_regional}
#| tbl-cap: Sidak-adjusted $p$-values for the regional one-sample problems.

one_sample_regional_tbl <- vnt |>
  select(region, region_name, t_max) |>
  summarise(t_max = list(c(t_max)), .by = c(region, region_name)) |>
  rowwise() |>
  mutate(res = list(one_sample(t_max))) |>
  unnest(res) |>
  select(region, Name, `p-value`) |>
  pivot_wider(names_from = Name, values_from = `p-value`)

multiple_testing <- function(tbl, m) {
  mutate(tbl, across(where(is.numeric), ~ 1 - (1 - .x)^m))
}

kable_format(multiple_testing(one_sample_regional_tbl,
  m = nrow(one_sample_regional_tbl)
))
```

## Pairwise comparisons

When the test of equality of the group mean densities is significant,
conducting a post-hoc analysis of pairwise group mean comparisons can
help identify which pairs differ. Testing whether the mean density in
group $g_1$ is equal to the mean density in group $g_2$ is then
equivalent to test whether the negative perturbation between these two
means is equal to the neutral element in the space $B^2(a,b)$, which is
the uniform density on $(a,b).$ In $L^2_0(a,b)$ it is also equivalent to
test whether the clr of the mean density of group $g_1$ is equal to the
clr of the mean density of group $g_2.$ Mathematically, these tests are
the same as in the previous section for the case of two groups. A
multiple testing correction is necessary when performing theses tests
for all pairs of groups.

### Test hypotheses

For comparing group $g$ and $g'$:

$H_0: \quad {\mathbb E}^B (\pi_{gi}) = {\mathbb E}^B (\pi_{g'i})$
equivalent to
$H_0: \forall g \quad {\mathbb E} (\text{clr }\pi_{gi}) = {\mathbb E} (\text{clr }\pi_{g'i})$

### Test statistics

Same as in previous section ? two ways: do a 2 group anova or do a one
sample test on the difference between two groups ? equivalent for
univariate but not here see Zhang page 137 in book. voir page 140 pour
la comparaison.

### Application: regional changes in temperature distribution over the period 1987-2016

```{r tbl-slopes_danova_pairwise}
#| tbl-cap: Sidak-adjusted $p$-values for pairwise comparisons.
vntp <- t(combn(c("NMM", "RRD", "NCC", "CHR", "SR", "MDR"), m = 2))
colnames(vntp) <- c("region1", "region2")

pairwise_anova <- function(region1, region2) {
  data <- vnte[vnt_province_region[colnames(vnte)] %in% c(region1, region2)]

  res <- fanova.tests(data, vnt_province_region[colnames(data)],
    test = stat_list,
    params = list(
      paramCS = 100,
      paramFmaxb = 100,
      paramTRP = list(B.TRP = 100)
    )
  )

  res[as_vector(stat_list)] |>
    sapply(function(x) c(x$pval)) |>
    c()
}

multiple_testing <- function(tbl, m) {
  mutate(tbl, across(where(is.numeric), ~ 1 - (1 - .x)^m))
}

pairwise_tbl <- as_tibble(vntp) |>
  rowwise() |>
  mutate(res = map2(region1, region2, pairwise_anova)) |>
  unnest_wider(res)

kable_format(multiple_testing(pairwise_tbl, m = nrow(pairwise_tbl)))
```

```{mermaid}
%%| label: fig-slopes_danova_pairwise_diagram
%%| fig-cap: Summary of pairwise tests.
flowchart TB
  NMM([NMM]) ~~~ MDR([MDR])
  NMM <-.-> SR([SR])
  NMM <--> CHR([CHR]) <--> SR
  NMM <-.-> RRD([RRD]) <-.-> CHR
  RRD <-.-> NCC
  NCC <--> CHR
  NMM <--> NCC([NCC]) <--> SR
  SR <-.-> MDR
  NMM ~~~ MDR([MDR])

style NMM fill:#440053,color:#fff,stroke:#000
style RRD fill:#414487,color:#fff,stroke:#000
style NCC fill:#2B778E,stroke:#000
style CHR fill:#22A885,stroke:#000
style SR fill:#7AD051,stroke:#000
style MDR fill:#FDE624,stroke:#000
```
