---
title: Testing mean densities with an application to climate change in Vietnam
date: today
date-format: long
license: CC BY
author:
  - name:
      given: Camille
      family: Mondon
    url: https://camillemondon.com
    email: camille.mondon@tse-fr.eu
    orcid: 0009-0007-4569-990X
    roles:
      - conceptualization
      - investigation
      - writing – original draft
    affiliations:
      - id: tse
        name: Toulouse School of Economics
        department: Mathematics and Statistics
        address: 1, Esplanade de l'Université
        city: Toulouse
        region: Occitanie
        country: France
        postal-code: 31000
        url: https://www.tse-fr.eu/department-mathematics-and-statistics
  - name:
      given: Huong Thi
      family: Trinh
    url: https://sites.google.com/tmu.edu.vn/huongtrinhthi/home
    email: trinhthihuong@tmu.edu.vn
    orcid: 0000-0002-5615-5787
    roles:
      - conceptualization
      - investigation
      - writing – original draft
    affiliations:
      - id: thuongmai
        name: Thuongmai University
        department: Faculty of Mathematical Economics
        city: Hanoi
        country: Vietnam
        url: https://tmu.edu.vn/
  - name:
      given: Josep Antoni
      family: Martín-Fernández
    email: josepantoni.martin@udg.edu
    orcid: 0000-0003-2366-1592
    roles:
      - supervision
      - conceptualization
      - investigation
      - writing – original draft
    affiliations:
      - id: girona 
        name: University of Girona
        department: Computer Science, Applied Mathematics and Statistics
        city: Girona
        region: Catalonia
        country: Spain
        url: http://www.udg.edu/
  - name:
      given: Christine
      family: Thomas-Agnan
    url: https://www.tse-fr.eu/fr/people/christine-thomas-agnan
    email: christine.thomas@tse-fr.eu
    orcid: 0000-0002-6430-3110
    roles:
      - supervision
      - conceptualization
      - investigation
      - writing – original draft
    affiliations:
      - ref: tse
abstract: |
  Given samples of density functions on an interval $(a,b)$ of ${\mathbb R}$, categorized according to a factor variable, we aim to test the equality of their mean functions both overall and across the groups defined by the factor.
  While the Functional Analysis of Variance (FANOVA) methodology is well-established for functional data, its adaptation to density functions (DANOVA) is necessary due to their inherent constraints of positivity and unit integral.
  To accommodate these constraints, we naturally use Bayes spaces methodology by mapping the densities using the centered log-ratio transformation into the $L^2_0(a,b)$ space where we can use FANOVA techniques.
  Many traditional contrasts in FANOVA rely on squared differences and can be reinterpreted as squared distances between Bayes perturbations within the densities space.
  We illustrate our methodology on a dataset comprising daily maximum temperatures across Vietnamese provinces between 1987 and 2016.
  Within the context of climate change, we first investigate the existence of a non-zero temporal trend of the densities of daily maximum temperature over Vietnam and then examine whether there is any regional effect on these trends. Finally, we explore odds ratio based interpretations allowing to describe the trends more locally.
keywords:
  - Bayes spaces
  - Analysis of variance
  - Density data
  - Functional data
  - Odds ratio
  - Bayes spaces
  - Climate change
  - Vietnam
citation:
  type: article-journal
  container-title: Stochastic Environmental Research and Risk Assessment
---

# Introduction

Climate change remains a central focus of scientific inquiry, as its
effects on weather patterns, ecosystems, and human livelihoods become
increasingly pronounced [@hultgren_estimating_2022]. According to the
World Meteorological Organization, the global mean temperature during
the 2011--2020 period has been $1.10 \pm 0.12^\circ\text C$ higher than
the average recorded between 1850 and 1900. Notably, weather-related
events accounted for nearly 94% of all recorded disaster-induced
displacements over the past decade
[@world_meteorological_organization_global_2023]. Among the various
global climate indicators, near-surface temperature is particularly
critical, as it directly influences human well-being and daily
activities [@gubler_statistical_2023; @world_meteorological_organization_state_2024; @schlenker_nonlinear_2009].
Temperatures can be measured using various indicators, such as heat
stress, growing degree days, killing degree days, temperature intervals,
or the entire temperature distribution
[@roberts_agronomic_2013; @vo_adaptation_2022; @espagne_impacts_2019; @schlenker_nonlinear_2009; @hultgren_estimating_2022].
Indeed, considering the entire temperature distribution captures the
full range of temperatures throughout each day and over multiple days,
offering a more comprehensive and informative perspective
[@hsiang_estimating_2017; @{trinh_scalar--distribution_2023}]. Vietnam
lies closer to the tropics than the equator and is significantly
influenced by the East Sea, resulting in a predominantly tropical
monsoon climate. The country's economy is heavily reliant on
agriculture, supported by its fertile deltas, mountainous regions, and
extensive coastline. Vietnam has been identified as one of the five
countries most vulnerable to the impacts of climate change
[@world_bank_group_climate_2021; @trinh_impact_2021]. Since 1960, the
country's mean annual temperature has risen by approximately 0.5°C to
0.7°C, with an estimated warming rate of 0.26°C per decade between 1971
and 2010 [@world_bank_group_climate_2021]. Consequently, as in many
other nations, analyzing climate change in Vietnam is of particular
interest due to its significant implications for agriculture, especially
rice cultivation [@tran_climate_2021; @trinh_impact_2021; @trinh_impact_2018].

With the increasing volume of recorded data, data science has seen the
rise of new types of observations like functional data. Functional data
analysis is currently a very active field of statistics, see
[@aneiros_functional_2022]. Of particular interest for our application
are density-valued data, which can be found in other areas of social
sciences (for example age distributions, income distributions,
expenditure distributions), and in other fields (for example particle
size distributions). Density data require a specific treatment within
the framework of functional data analysis, due to their inherent
constraints, namely non-negativity and integration to one. It is
typically assumed that each observation corresponds to a sampled
continuous density function, belonging to an infinite-dimensional
function space, as discussed in @petersen_modeling_2022. A smoothing
tool is necessary to fill the gap between the discrete data and the
continuous temperature density objects and this procedure step is called
preprocessing. Among the methods described in @petersen_modeling_2022,
we use the Bayes spaces approach first introduced in
@egozcue_hilbert_2006 and later on developed in
[@van_den_boogaart_bayes_2010; @van_den_boogaart_bayes_2014].

In this paper, we focus on testing mean density functions, investigating
their equality to a reference density as well as their equality across
groups. Since density functions share similar constraints with
compositional vectors, albeit in a continuous form, we build upon
techniques from compositional data analysis (CoDA) but also from the
functional analysis of variance (FANOVA) framework, introduced in
[@ramsay_functional_2005; @kokoszka_introduction_2017].

Before comparing group means, a first question can be to find whether
the expected density is equal to a given reference, for example the
uniform density. We adapt the one-sample test from the functional
framework [see @zhang_analysis_2013] to density functions in @sec-one-sample.

As outlined by @martin-fernandez_interpretation_2015, the analysis of
grouped data typically begins with testing the equality of group means
and a widely used approach for this purpose in CoDA is the multivariate
analysis of variance (MANOVA) contrast, which needs to be adapted here
to continuous objects. On the other hand, the FANOVA techniques
(reminded in @sec-reminders_fanova) must be adapted when applied to
density functions because they reside in the constrained space
${B}^2(a,b)$. We present five test statistics in @sec-anova for the
problem of testing the equality of group means of density functions,
which we call DANOVA for distributional analysis of variance.

When an ANOVA test leads to a rejection of the null hypothesis of equal
group means, a related question of interest is the pairwise comparison
of group means [see for example @martin-fernandez_interpretation_2015
who present additional techniques for interpreting the differences
between the groups in CoDA]. This question is treated in @sec-pairwise.

Finally, in @sec-intervalwise, to go beyond the global comparisons of
densities and do a more local analysis, we adapt a technique based on
odds ratios from @maier_additive_2025 to infer the relative mass of
the densities over specific intervals.

We apply the above tools to address different questions relative to
climate change in Vietnam.

We use the distributions of maximum temperatures in the provinces of
Vietnam to compare the six administrative regions in terms of climate
change. The dataset and its preprocessing are presented in @sec-data. After
constructing a trend slope density for each province summarizing its
time evolution, we investigate maximum temperature distribution changes
through these trend slope densities sample. Then, an analysis of
variance of the slope densities in @sec-reg_dist_anova detects the pairs of regions which
differ the most in terms of evolution of the temperature distribution.
Finally in @sec-regional_interpretation, we compare the relative
frequencies in different subintervals of the trend distribution allowing
a more localized analysis of temperature density evolution.

# Reminders {#sec-reminders}

## Reminders on functional one-sample problem {#sec-reminders_1sample}

In Chapter 5 of @zhang_analysis_2013, the functional sample
$(f_1, \dots, f_n),$ is assumed to arise from the following model
$$f_{i}(x)= f_0(x) + v_{i}(x),$$ {#eq-fun_mod_1sample}
where $f_0(x) = {\mathbb E}(f_{i})(x)$ is the unknown mean function and the
stochastic error process $v_{i}$ has mean $0$.

Two test statistics are available: the $L^2$-norm test statistic
[@zhang_analysis_2013] and the PC approach
[@kokoszka_introduction_2017].

## Reminders on functional analysis of variance {#sec-reminders_fanova}

In the classical framework [@zhang_analysis_2013, p. 144] of functional
analysis of variance, or simply FANOVA, we observe $G$ independent
functional samples, denoted by $(f_{g1}, \ldots, f_{gn_g})$ for
$1 \leq g \leq G$, from stochastic processes with values in $L^2(a,b)$,
satisfying for $1 \leq i \leq n_g$ and $a \leq x \leq b$
$$f_{gj}(x)= f_g(x) + v_{gj}(x),$$ {#eq-fun_mod}
where $f_g(x) = {\mathbb E}(f_{gj})(x)$ is the unknown mean function in group
$g$ and the stochastic error process $v_{gj}$ has mean $0$ and common
covariance operator. The total sample size is $n=\sum_{g=1}^G n_g$. The
overall sample mean curve $\bar f_{..}$ and the group sample mean curves
$\bar f_{g.}$ are respectively defined by
$$\begin{aligned}
\bar f_{..}(x) &= \frac{1}{n}\sum_{g=1}^G \sum_{i=1}^{n_g} f_{gj}(x) \\
\bar f_{g.}(x) &= \frac{1}{n_g}\sum_{i=1}^{n_g} f_{gj}(x).
\end{aligned}$$
The pointwise between-group mean square error and the
pointwise within-group mean square error at $x$ are respectively defined
by
$$\operatorname{SSB}(x) = \sum_{g=1}^G n_g \left( \bar f_{g.}(x) - \bar f_{..}(x) \right)^2$$ {#eq-SSBloc}
and
$$\operatorname{SSW}(x) = \sum_{g=1}^G \sum_{i=1}^{n_g} \left( f_{gj}(x)-\bar f_{g.}(x) \right)^2.$$ {#eq-SSWloc}
@ramsay_functional_2005 extend the classical F-test
and propose the pointwise functional F-ratio to test the equality of the
group mean curves at a given point $x$, using the local F-statistic
$$F(x)=\frac{\operatorname{SSB}(x)}{\operatorname{SSW}(x)}.$$ {#eq-locF}
For a global assessment of the equality of the group mean curves on the whole
interval of variation of their argument, @zhang_statistical_2007 and
@zhang_analysis_2013 introduce $L^2$-norm based tests as well as
F-type test statistics. Later on, @zhang_one-way_2014 propose the GPF
test based on the integral of the pointwise F-ratio statistic (see @sec-anova)
and the $F_\text{max}$ test based on the maximum of the pointwise
F-test. It is then necessary to approximate the null distribution of
these statistics. This can be achieved using a permutation based
procedure as in @ramsay_functional_2005 or a bootstrap procedure as in @zhang_new_2019.

## Reminders on distributional data analysis and Bayes spaces {#sec-remind_dda}

@egozcue_hilbert_2006 and
@van_den_boogaart_bayes_2010 [-@van_den_boogaart_bayes_2014] define the
Bayes spaces of probability density functions relative to a reference
measure $\lambda$ on an interval of $\mathbb R$, in a similar fashion as
$L^2(\lambda)$ spaces. In the following, the measure $\lambda$ will be
the Lebesgue measure on a finite interval $[a,b]$ and we will denote
these spaces by ${B}^2(a,b)$. We consider the separable Hilbert space
$L^2_0(a,b)$ of square-integrable functions with a zero integral on
$(a,b)$ equipped with the inner product
$\langle f,g \rangle_{L^2} = \int_a^b fg \,\mathrm d\lambda$. For any
measurable function $\varphi : [a,b] \rightarrow \mathbb R$ that is
positive almost everywhere and such that the function $\log (\varphi)$
is integrable, we can define its centered log-ratio transform
$\operatorname{clr}(\varphi)$ $$
  x \in [a,b] \mapsto \log(\varphi (x)) - \frac1{b-a} \int_a^b \log(\varphi)(u) \mathrm du.$$ {#eq-clr}
Note that by construction $\operatorname{clr} (\varphi) \in L^2_0(a,b).$
Conversely for each $f \in L^2_0 (\lambda)$, the equivalence class
$\operatorname{clr}^{-1} (f) = \{ \alpha \exp (f), \alpha >0 \}$
contains positive functions that are equal almost everywhere, up to a
multiplicative constant. Among them there is a unique probability
density function $\varphi$ (thus satisfying
$\int_a^b \varphi(u) \, \mathrm du = 1$) that we use to represent
$\operatorname{clr}^{-1} (f)$. For any element $p$ of
$\operatorname{clr}^{-1} (f),$ the closure of $p,$ denoted by
${\cal C}(p)$, is defined to be $\varphi$.

Then the Bayes Hilbert space with Lebesgue reference measure on the
interval $[a,b]$ is the set of probability density functions
$${B}^2 (a,b) = \left\{ \operatorname{clr}^{-1} (f), f \in L^2_0 (a,b) \right\}$$
equipped with the only separable Hilbert space structure
$(\oplus, \odot, \langle \cdot, \cdot \rangle_{B^2})$ that makes the
centered log-ratio transform an isometry between
$({B}^2 (a,b), \oplus, \odot, \langle \cdot, \cdot \rangle_{{B}^2})$ and
$(L^2_0 (a,b), +, \cdot, \langle \cdot, \cdot \rangle_{L^2})$. The
resulting addition $\oplus$ is called Aitchison perturbation ($\ominus$
denoting the negative perturbation) and the scalar multiplication
$\odot$ is called Aitchison powering.

Since it is important to understand some further developments, we recall
that, given two elements $\varphi_1$ and $\varphi_2$ of the Bayes space,
the perturbation density $\varphi_2 (x) \ominus \varphi_1 (x)$ is the
closure of the ratio $\frac{\varphi_2 (x)}{\varphi_1 (x)}$, which is equal to
$\frac{{\varphi_2 (x)}/{\varphi_1 (x)}}{\int_a^b {\varphi_2 (u)}/{\varphi_1 (u)}du}$.

Following @van_den_boogaart_bayes_2014, it is also possible to use the
centered log-ratio transform in order to transport the Borel sets to
${B}^2 (a,b)$, as well as the expectation and covariance of
$L^2_0 (a,b)$-valued random variables. For a random density $\varphi$ in
${B}^2(a,b)$, one can define the expected value in the Bayes space as:
$$
\mathbb{E}^B(\varphi) = \operatorname{clr}^{-1} (\mathbb E [\operatorname{clr}(\varphi)])$$ {#eq-expB}
and the covariance operator, for $\psi \in B^2(a,b)$, by:
$$\begin{split}
\operatorname{Cov}^B [\varphi] (\psi) &= \operatorname{clr}^{-1} (\operatorname{Cov} [\operatorname{clr} \varphi] (\operatorname{clr} \psi)) \\
&= \mathbb{E}^B \left[ \langle\varphi \ominus \mathbb E ^B (\varphi), \psi \rangle_{B^2} \odot \left( \varphi \ominus \mathbb E^B (\varphi) \right) \right].
\end{split}$$ {#eq-covB}

# Testing and interpreting mean densities in the Bayes space {#sec-global}

All testing procedures in this section involve hypotheses about the
global behavior of mean densities in the whole interval $[a,b].$ In the
application, the global tests of this section will be applied to the
samples of trend slope densities in order to investigate the regional
effect on climate change.

## One-sample problem {#sec-one-sample}

In this section, we consider a sample of densities
$\varphi_i, i=1, \ldots n$ in the Bayes space $B^2(a,b)$ distributed as
$\varphi$. We are given a reference density $\varphi_0$ of particular
interest and we wish to test the equality to $\varphi_0$ of the mean
density $\mathbb{E}^B(\varphi)$:
$$H_0: {\mathbb E}^B (\varphi) = \varphi_0.$$
Since
$$\begin{split}
{\mathbb E}^B (\varphi) = \varphi_0
&\Leftrightarrow \operatorname{clr} {\mathbb E}^B (\varphi) = \operatorname{clr} \varphi_0 \\
&\Leftrightarrow {\mathbb E} (\operatorname{clr} \varphi) = \operatorname{clr} \varphi_0,
\end{split}$$
the global one-sample test on mean densities is equivalent
to a one sample test in the $L^2(a,b)$ space on the clr transformed
densities $f_i = \operatorname{clr} (\varphi_i)$ for which we may apply
a classical one sample test for functional data as described in Chapter
5 of @zhang_analysis_2013. As in @zhang_analysis_2013, the
functional sample $(f_1, \dots, f_n),$ is assumed to arise from model @eq-fun_mod_1sample.

Two test statistics are available: the $L^2$-norm test statistic
[@zhang_analysis_2013] and the PC approach
[@kokoszka_introduction_2017]. We now adapt them to our density
framework.

#### $L^2$-norm approach {#l2-norm-approach .unnumbered}

We consider the following statistic defined in the Bayes space by
$$\begin{split}
T_\text{norm} &= n \| \bar{\varphi}_{.} - \varphi_0 \|^2_{B^2(a,b)} \\
&= n \| \operatorname{clr} (\bar{\varphi}_{.}) - \operatorname{clr} (\varphi_0) \|^2_{L^2_0(a,b)},
\end{split}$$
where $\bar{\varphi}_{.}$ denotes the sample mean of the
densities in $B^2(a,b)$. We can see that it does correspond to an
$L^2$-norm statistic to test the equality in $L^2_0$ of the clr
transform $f_i= \operatorname{clr} (\varphi_i)$ of our sample of
densities $\varphi_i$. We can therefore apply the classical one sample
test for functional data described in Chapter 5 of @zhang_analysis_2013.

The asymptotic distribution of this statistic under the null is that of
a linear combination of chi-squared statistics weighted by the
eigenvalues of the covariance operator, see
[@kokoszka_introduction_2017]. For the computation of the $p$-value of
the test, the eigenvalues have to be estimated using FPCA and
@kokoszka_introduction_2017 provide code for the computation of the
$L^2$-norm.

#### PC approach {#pc-approach .unnumbered}

To eliminate the dependence of the limiting distribution upon the
unknown eigenvalues, @kokoszka_introduction_2017 propose to truncate
the Functional Principal Component Analysis (FPCA) and use a
Hotelling-type test statistic. It turns out that this statistic applied
to the clr-transformed densities coincides with that obtained with the
simplicial FPCA in $B^2(a,b)$ from @hron_simplicial_2016. Keeping the
first $p$ terms, the statistic $T_{\operatorname{PC}}$ is given by
$$\begin{split}
T_{\operatorname{PC}} &= n\sum_{k=1}^p \frac{\langle \operatorname{clr} (\bar{\varphi}_{.}) - \operatorname{clr} (\varphi_0), \hat{v}_k \rangle_{L^2_0(a,b)}^2}{\hat \lambda_k} \\
&= n\sum_{k=1}^p \frac{\langle \bar{\varphi}_{.} \ominus \varphi_0, \operatorname{clr}^{-1}(\hat{v}_k) \rangle_{B^2(a,b)}^2}{\hat \lambda_k}.
\end{split}$$
where $\hat\lambda_k$ are the estimated non-null
eigenvalues and $\hat v_k$ the corresponding estimated eigenfunctions in
$L^2_0(a,b)$ from the FPCA. The limiting distribution of
$T_{\operatorname{PC}}$ under the null is then a simple chi-squared with
$p$ degrees of freedom. @kokoszka_introduction_2017 recommend using
the smallest value of $p$ for which the explained variance exceeds
$85 \%$, and to use the norm approach otherwise.

## Distributional ANOVA {#sec-anova}

We observe $G$ independent samples of densities, denoted by
$(\varphi_{g1}, \ldots, \varphi_{gn_g})$ for $1 \leq g \leq G$, from
stochastic processes with values in $B^2(a,b)$, satisfying for
$1 \leq i \leq n_g$ and $a \leq x \leq b$ 
$$\varphi_{gj}(x)= (\varphi_g \oplus u_{gj})(x),$$ {#eq-danova_model}
where $\varphi_g(x) = {\mathbb E}^B(\varphi_{gj})(x)$ is the unknown mean
density in group $g$ and the stochastic error process $u_{gj}$ has mean
equal to the uniform distribution on $(a,b)$ and common covariance
operator (defined by @eq-covB in Bayes spaces). The total sample size is
$n=\sum_{g=1}^G n_g$. As above the overall sample mean density is
defined as
$$\bar \varphi_{..} (x) = \frac1n \odot \bigoplus_{g=1}^G \bigoplus_{i=1}^{n_g} \varphi_{gj} (x)$$
and the sample mean density in group $g$ as
$$\bar\varphi_{g.} (x) = \frac1{n_g} \odot \bigoplus_{i=1}^{n_g} \varphi_{gj} (x).$$
Applying the FANOVA formulas to the clr-transformed densities, we adapt
the FANOVA framework to densities and define the pointwise between-group
mean square error and the pointwise within-group mean square error at
$x$ by
$$\operatorname{SSB}(x) = \sum_{g=1}^G {n_g} (\operatorname{clr}(\bar\varphi_{g.})(x) - \operatorname{clr}(\bar\varphi_{..})(x))^2
$$ {#eq-SSBlocBayes}
and
$$\operatorname{SSW}(x) = \sum_{g=1}^G \sum_{i=1}^{n_g} (\operatorname{clr}(\varphi_{gj})(x) - \operatorname{clr}(\bar\varphi_{g.}) (x))^2
$$ {#eq-SSWlocBayes}

Using the notations of @sec-remind_dda,
the objective of DANOVA is to test the null hypothesis $H_0$ that the
$G$ group mean densities ${\mathbb E}^B(\varphi_g)$ are equal against
the usual alternative that at least two means are different, where mean
density here is understood in the Bayes space sense:
$$H_0: \forall g, \quad {\mathbb E}^B (\varphi_{gj}) = {\mathbb E}^B (\varphi_{1j}).$$ {#eq-Hzero}
Let us first note that for two densities $\varphi_1$ and $\varphi_2$ we
have the following equivalence
$${\mathbb E}^B (\varphi_1) = {\mathbb E}^B (\varphi_{2}) \Leftrightarrow {\mathbb E} (\operatorname{clr }\varphi_{1}) = {\mathbb E} (\operatorname{clr }\varphi_{2}),$$ {#eq-equivalence}
since the clr of the expected density in the Bayes sense is equal to the
expected value in the classical sense of the clr of the density in the
corresponding $L^2_0$ space. Therefore, the assumption @eq-Hzero
is equivalent to
$$H_0: \forall g, \quad {\mathbb E} (\operatorname{clr}\varphi_{gj}) = {\mathbb E} (\operatorname{clr}\varphi_{1j}).$$
In order to adapt FANOVA to DANOVA, we can thus simply apply the FANOVA
techniques to the clr transformed densities in the functional space
$L^2_0.$ The [R]{.sans-serif} package `fdANOVA` [described in @gorecki_fdanova_2019] provides several FANOVA options and we focus on
the following five test statistics. Their evaluation requires the
computation of the following quantities: the pointwise variations
$\operatorname{SSB}(x)$ and $\operatorname{SSW}(x)$ from @eq-SSBlocBayes and
@eq-SSWlocBayes, and pairwise negative perturbations between
group means for the statistic proposed by @cuevas_anova_2004. Let us
briefly describe the five test statistics and try to provide when
possible a Bayes space expression.

#### $\operatorname{L^2B}$ statistic {#operatornamel2b-statistic .unnumbered}

We adapt the $L^2$-norm based test statistic given in the framework of
FANOVA [@zhang_statistical_2007] to the context of densities:
$$\operatorname{L^2B} = \int_a^b \operatorname{SSB}(x)dx,$$ {#eq-zhang}
where $\operatorname{SSB}(x)$ is given by
@eq-SSBlocBayes. Note that an alternative formula for
$\operatorname{L^2B}$ can be written using the norm of the Bayes space:
$$\operatorname{L^2B} = \sum_{g=1}^G n_g \| \bar\varphi_{g.} \ominus \bar\varphi_{..}\|^2_{B^2},$$ {#eq-zhang2}
which shows that this statistic is invariant under almost-everywhere
equality. In the @gorecki_fdanova_2019 `fdANOVA` `R`
package, the `L2B` implementation uses an asymptotic distribution of the
test statistic.

#### $\operatorname{FB}$ statistic {#operatornamefb-statistic .unnumbered}

The F-type tests from @shen_f_2004 and @zhang_statistical_2011 use
both within and between variation:
$$\begin{split}
\operatorname{FB} &= \frac{{\int_a^b \operatorname{SSB}(x)dx}/(G-1)}{\int_a^b \operatorname{SSW}(x)dx /(n-G)} \\
&= \frac{\sum_{g=1}^G n_g \| \bar\varphi_{g.} \ominus \bar\varphi_{..}\|^2_{B^2}/(G-1)}{\sum_{g=1}^G \sum_{i=1}^{n_g} \| \varphi_{gj} \ominus \bar\varphi_{g.}\|^2_{B^2}/(n-G)}.
\end{split}$$ {#eq-Fvdb}
We choose the bias-reduced `FB` implementation for the
computation of the $p$-value. Under the null hypothesis, this statistic
is asymptotically equal to a linear combination of independent $\chi^2$
for the numerator and the denominator, and can thus be approximated by a
Fisher distribution [@zhang_statistical_2011, Theorem 1 and equation
(2.16)]. The test statistic coincides with that of
@van_den_boogaart_bayes_2014 by the second equality in @eq-Fvdb.
However, @van_den_boogaart_bayes_2014 use a bootstrap procedure rather
than the above approximation.

#### $\operatorname{CS}$ statistic {#operatornamecs-statistic .unnumbered}

The CS statistic from @cuevas_anova_2004 uses pairwise differences.
When applied to clr of densities, due to the linearity of
$\operatorname{clr}$, we get
$$\begin{split}
\operatorname{CS} &= \sum_{g<g'} n_g \int_a^b \left( \operatorname{clr} {\bar\varphi_{g.}}(x)- \operatorname{clr} {\bar\varphi_{g'.}}(x) \right)^2dx \\
&= \sum_{g<g'} n_g \| \bar\varphi_{g.} \ominus \bar\varphi_{g'.} \|^2_{B^2}.
\end{split}$$ {#eq-cuevas}
The `CS` implementation in the `fdANOVA` package uses a
heteroscedastic assumption and parametric bootstrap. Note that an
alternative formula for CS is the sum of the Bayes norms of all pairwise
differences between groups.

#### $\operatorname{GPF}$ statistic {#operatornamegpf-statistic .unnumbered}

The GPF statistic from @zhang_one-way_2014 integrates the pointwise
F-ratio instead of integrating separately the pointwise within and
between variations:
$$\operatorname{GPF} = \int_a^b \frac{\operatorname{SSB}(x)/(G-1)}{\operatorname{SSW}(x)/(n-G)}dx.$$ {#eq-GPF}
Under the null hypothesis, this statistic is asymptotically equal to a
linear combination of independent $\chi^2$, which can be approximated by
a $\chi^2$-distribution [@zhang_one-way_2014, Proposition 1]. In the
implementation of the `fdANOVA` package, `GPF` is divided by $b-a$, and
the null distribution is modified accordingly. @gorecki_fdanova_2018
(page 5) claim that the GPF test is more powerful than the F-type tests.

#### $F_{\max}$ statistic {#f_max-statistic .unnumbered}

The $F_{\max}$ statistic from @zhang_new_2019 rather computes the
supremum of the pointwise F-ratio instead of integrating it as in `GPF`:
$$F_{\max} = \sup_{x \in (a,b)} \frac{\operatorname{SSB}(x)/(G-1)}{\operatorname{SSW}(x)/(n-G)}.$$ {#eq-Fmax}
We consider the implementation `Fmaxb` which bootstraps the distribution
under the null hypothesis. The statistic $F_{\max}$ is the only one that
is not invariant under almost-everywhere equality: changing one value of
a density in the dataset might change the value of $F_{\max}$.

Finally, note that the statistics GPF and $F_{\max}$ cannot be written
straightforwardly in terms of norms in the Bayes space.
@gorecki_fdanova_2018 compute empirical powers of some FANOVA tests
and conclude that there is no method that performs best in all
situations.

## Pairwise comparisons {#sec-pairwise}

When the test of equality of group mean densities is significant,
conducting a post-hoc analysis of pairwise group mean comparisons can
help identify which pairs differ. The null hypothesis for comparing
groups $g$ and $g'$ is:
$$H_0: \quad {\mathbb E}^B (\varphi_{gj}) = {\mathbb E}^B (\varphi_{g'j}),$$
which is equivalent to
$$H_0: \quad {\mathbb E} (\operatorname{clr}\varphi_{gj}) = {\mathbb E} (\operatorname{clr}\varphi_{g'j}).$$

Testing whether the mean density in group $g_1$ is equal to the mean
density in group $g_2$ can be viewed as a two-sample ANOVA test. A
multiple testing correction is necessary when performing these tests for
all pairs of groups.

## Interval-wise interpretation {#sec-intervalwise}

The global tests presented in the previous section, whether one-sample
tests or analysis of variance, enable us to draw inferences about
differences in the behavior of mean densities across the entire range of
the underlying random variable. Now, we aim to make more localized
statements regarding specific intervals of this range.

Unfortunately, the idea of adapting local tests from FANOVA does not
work for two reasons. First of all, the meaning of a test of the
equality of two mean densities evaluated at a given point $x$ is unclear
since densities are only defined almost everywhere, unless we impose
continuity restrictions. The second reason is that the equivalence
@eq-equivalence between the equality of two mean densities
in a Bayes space and the equality of their clr transform is not valid
anymore at the local level because the clr transform evaluated at $x$
involves all values of the log density and not only its value at $x$.

For these reasons, we now turn attention to some interpretation tools
presented in @maier_additive_2024 based on odds ratios.

For some subintervals $I \subset [a,b]$ we wish to compare the relative
probability to be in $I$ versus its complement in $(a,b)$ between two
densities $\varphi_1$ and $\varphi_2$. Let us start with an
infinitesimal comparison that only implies any two points $x,z$ in
$(a,b)$.

Given $\varphi_1$ and $\varphi_2$, @maier_additive_2024 [pp. 10 and 11,
Proposition 3.1] define the infinitesimal odds ratios
$\operatorname{OR}_{x\mid z}$ of $\varphi_2$ versus $\varphi_1$ as
$$\operatorname{OR}_{x\mid z} = \frac{\frac{{\varphi}_2(x)}{{ \varphi}_2(z)}}{\frac{{\varphi}_1(x)}{{ \varphi}_1 (z)}}.$$ {#eq-ORdef1}
$\operatorname{OR}_{x\mid z}$ is the ratio of the odds of $x$ versus $z$
for $\varphi_2$ by the odds of $x$ versus $z$ for $\varphi_1$. Let us
prove that this quantity coincides with the relative density of the
negative perturbation $\beta = \varphi_2 \ominus \varphi_1$ at the two
points $x$ and $z$ in $(a,b)$. Indeed, we can first
rewrite the double ratio differently with $x$ in the top ratio and $z$
in the bottom one and then use the fact that the integral of the
numerator is the same as that of the denominator:
$$\begin{split}
\operatorname{OR}_{x\mid z}
&= \frac{\frac{\varphi_2(x)}{\varphi_1(x)}}{ \frac{\varphi_2(z)}{ \varphi_1 (z)}}
= \frac{{\cal C}(\frac{\varphi_2(x)}{\varphi_1(x)})}{{\cal C}(\frac{\varphi_2(z)}{\varphi_1(z)})} \\
&= \frac{{\varphi}_2(x) \ominus { \varphi}_1(x)}{{ \varphi}_2(z) \ominus{ \varphi}_1(z)}
= \frac{\beta (x)}{\beta (z)}.
\end{split}$$ {#eq-ORdef2}
where $\cal C$ denotes the closure operator whose
definition is recalled in @sec-remind_dda.
This justifies that we write the infinitesimal odds ratios between two
densities $\varphi_1$ and $\varphi_2$ as a function
$\operatorname{OR}_{x\mid z} (\beta)$ of their negative perturbation
$\beta$.

Let us now use these infinitesimal odds ratios to compare the densities
on subintervals of $[a,b]$. Using part (a) of Proposition 3.1 in
@maier_additive_2024, which states that if
$\operatorname{OR}_{x\mid z} (\beta) > 1$ for all $x,z$ when $x$ is in a
given interval $A$ and $z$ in a given interval $B$, we may conclude that
conditional on the underlying variable being in $A$ or $B$, the odds
(according to the density ${\varphi}_2$) of being in $A$ are larger than
the odds (according to the density ${\varphi}_1$) of being in $A$.
Because the probability of an event is an increasing function of its
odds, the same is true for the corresponding probabilities so that going
from $\varphi_1$ to $\varphi_2$ there is a transfer of the probability
mass from $B$ to $A$.

Using the fact that $\operatorname{OR}_{x\mid z} (\beta) >1$ is
equivalent to $\beta(x) > \beta(z)$, we are going to show that the curve
of $\beta$ allows to draw conclusions about the relative size of the
probability mass of some chosen intervals according to $\varphi_1$ and
$\varphi_2$. For a given level $\tau = \beta(x_0) >0$, let the
collection of intervals (or unions of intervals) $A_{\tau}$ be defined
by $A_{\tau} = \{ x: \beta(x) > \tau \}$ and let $A_{\tau}^c$ be its
complement. Then we have for almost all $x \in A_{\tau}$ and almost all
$z \in A_{\tau}^c$
$$\operatorname{OR}_{x\mid z}(\beta) = \frac{\beta(x)}{\beta(x_0)}\frac{\beta(x_0)}{\beta(z)} > 1.$$ {#eq-atau}
Therefore, we may say that the probability to lie in $A_{\tau}$ under
the distribution of $\varphi_2$ is higher than that according to the
distribution of $\varphi_1$.

As we can see in @eq-ORdef2 this type of interpretation is useful as soon as
we are interested in a negative perturbation between two densities. In
particular, as we will see in the next section, we can use it to compare
in a local fashion two mean temperature densities in two regions or at
two different dates. It will also be useful to interpret regional trends
in temperature distributions which turn out to be negative perturbations
between the temperature density at two consecutive times. More
generally, provided we have a linear model for our density objects, the
same tool can be used to make marginal effects interpretations for a
scalar covariate parameter.


# Trends in yearly temperature distributions in Vietnam over the period 1987-2016 {#sec-data}

## Description

The climate data used in this study comprises daily temperatures and
precipitation, sourced from the Climate Prediction Center (CPC)
database, developed by the National Oceanic and Atmospheric
Administration (NOAA). This data provides historical records of the
maximum and minimum temperatures for a 0.5-degree by 0.5-degree grid of
latitude and longitude. For this study, we focus exclusively on maximum
temperatures as they are a direct indicator of heat stress on both
humans and the environment. The data spans the years 1987 to 2016. In
order to address the discrepancy between the geographical coordinates
and the municipalities of Vietnam (63 provinces),
@espagne_impacts_2019 performed a spatial interpolation step from the
gridded data to the provincial level.

This process results in 365 or 366 daily maximum temperature records for
each province and year.

During the study period, maximum temperatures (hereafter denoted
$T_\text{max}$) range from -5°C to nearly 45°C. However, extreme cold
and extreme heat are rare. Moreover subintervals with no observations
result in zero density value which is problematic for the logarithm. For
the purpose of our analysis, we focus on a temperature range of 12°C to
40°C. Temperatures below 12°C are adjusted upward to 12°C, while
temperatures exceeding 40°C are capped at 40°C. As we can see on @fig-map, there are
$n=63$ provinces in Vietnam grouped into $G=6$ socioeconomic regions. We
use the following acronyms for the regions: NMM for Northern Midlands
and Mountains region ($n_1=14$ provinces), RRD for Red River Delta
region ($n_2=12$ provinces), NCC for North Central Coast region
($n_3=13$ provinces), CHR for Central Highlands region ($n_4=5$
provinces), SR for Southeast region ($n_5=6$ provinces) and MDR for
Mekong Delta region ($n_6=13$ provinces). @trinh_impact_2021 highlight
the fact that the regions have their own weather peculiarities. Some
regions in the north have four seasons: winter, spring, summer, and
autumn, while some regions in the south experience two distinct seasons:
the dry season (November to April) and the rainy season (May to October)
[@trinh_impact_2021; @world_bank_group_climate_2021]. For example,
according to @trinh_impact_2021, during the period 1960-2000 winter
temperatures rose faster than those of the summer, and temperatures in
the northern zones rose faster than in the southern zones. The objective
of this application is first of all to determine whether there is a
regional factor in the evolution of temperature distributions over time.
The groups are thus the six regions of Vietnam. Secondly, we will
describe the changes in regional temperature distributions in some
subintervals of the temperature range.

```{r setup}
#| include: false
#| echo: false
#| purl: false
source("setup.R")
```

```{r fig-map}
#| fig-cap: Map of the provinces of Vietnam colored by region.
#| fig-height: 7
#| warning: false
#| message: false
library(boot)
library(CompQuadForm)
library(dda)
library(fdANOVA)
library(fda.usc)
library(MVN)
library(sf)
library(tidyverse)
library(colorspace)
set.seed(10983)

data(vietnam_temperature_dd)
data(vietnam_regions)
data(vietnam_provinces)

lat_order_code <- c("NMM", "RRD", "NCC", "CHR", "SR", "MDR")
lat_order <- c(
  "Northern Midlands and Mountains",
  "Red River Delta",
  "North Central Coast",
  "Central Highlands",
  "Southeast",
  "Mekong Delta"
)
vietnam_regions$name <- factor(str_wrap(vietnam_regions$name, 20), levels = str_wrap(lat_order, 20))
vietnam_regions$code <- factor(vietnam_regions$code, levels = lat_order_code)
vietnam_temperature_dd$region <- factor(vietnam_temperature_dd$region,
  levels = lat_order_code
)

vietnam_provinces |>
  left_join(vietnam_regions, by = c("region" = "code")) |>
  rename(region_name = name) |>
  ggplot(aes(fill = region_name)) +
  scale_fill_viridis_d(option = "magma", end = 0.8) +
  geom_sf() +
  scale_x_continuous(breaks = seq(from = -180, to = 180, by = 2)) +
  labs(x = "Longitude", y = "Latitude", fill = "Region")
```

## Smoothing

We will assume that the observed data arise from random densities,
themselves sampled from a hypothesized and unknown distribution on a
Bayes space. Since the elements of the Bayes spaces are density
functions, whereas the observed data (usually samples of real-valued
observations or histograms) is always discrete, a preprocessing step is
necessary to transform the observed samples or histograms into a sample
of probability density functions. We will follow the preprocessing
procedure of
[@machalova_preprocessing_2016; @machalova_compositional_2021] to
transform our yearly samples of maximum temperatures at the province
level into density functions that we assume to belong to a
finite-dimensional subspace $\mathcal C_q^\Delta (a,b)$ of the Bayes
Hilbert space $B^2(a,b)$, made of compositional splines. A compositional
spline (CB-spline) is a probability density function whose logarithm is
a polynomial spline.

The process starts by the choice of a basis of normalized B-spline
functions for the space $\mathcal S_d^\Delta (a,b) \subset L^2(a,b),$ of
polynomial splines of order $d$ (degree less than or equal to $d-1$) and
inside knots $\Delta = (\delta_1, \ldots, \delta_k)$, of dimension $d+k$
[see @schumaker_spline_1981 for a complete description]. To accommodate
the zero-integral constraint, @machalova_preprocessing_2016 introduce
the ZB-spline functions, denoted by $Z_\ell(x), 1 \leq \ell \leq d+k-1$
for the subspace
$\mathcal Z_d^\Delta (a,b) = \mathcal S_d^\Delta (a,b) \cap L^2_0(a,b)$
of dimension $d+k-1$, the loss of one dimension being due to the
zero-integral constraint. Finally the inverse clr of the ZB-spline basis
functions, called the CB-spline basis functions, denoted by
$C_\ell =\operatorname{clr}^{-1}(Z_\ell)$ generate the space
$\mathcal C_d^\Delta (a,b),$ a Euclidean subspace of ${B}^2 (a,b)$ of
dimension $k+d-1,$ made of compositional splines on $[a,b]$ of order $d$
with knots sequence $\Delta.$ The expansion @eq-expan
of a density $\pi_{gj}^{t}$ of the subspace $\mathcal C_d^\Delta (a,b)$
of ${B}^2(a,b)$ and the corresponding expansion @eq-expan2 of its clr transform in the corresponding
ZB-spline basis generating the space $\mathcal Z_d^\Delta (a,b)$ are
then given by:
$$\pi_{gj}^{t}(x) = \bigoplus_{\ell=1}^{d+k-1} \left[\pi_{gj}^{t}\right]^{C_\ell}\odot C_\ell(x)$$ {#eq-expan}
or equivalently
$$\operatorname{clr}(\pi_{gj}^{t})(x) = \sum_{\ell=1}^{d+k-1} \left[\pi_{gj}^{t}\right]^{C_\ell} Z_\ell(x),$$ {#eq-expan2}
where $\left[\pi_{gj}^{t}\right]^{C_\ell}$ is the
$\ell$-th coefficient of $\pi_{gj}^{t}$ in the CB-spline basis $C$. Note
that the coefficients are the same in the two equations. The
coefficients of the global sample mean and regional sample means, needed
to compute the pointwise between-groups and within-groups mean square
error, are readily obtained by
$$\begin{split}
\operatorname{clr} \bar{\pi}^{t}_{g.}(x) &= \frac{1}{n_g} \sum_{i=1}^{n_g} \operatorname{clr} \pi_{gj}^{t}(x) \\
&= \frac{1}{n_g} \sum_{i=1}^{n_g} \sum_{\ell=1}^{d+m-1} \left[\pi_{gj}^{t}\right]^{C_\ell} \cdot Z_\ell(x) \\
&= \sum_{\ell=1}^{d+m-1} \left( \frac{1}{n_g} \sum_{i=1}^{n_g}\left[\pi_{gj}^{t}\right]^{C_\ell} \right) \cdot Z_\ell(x),
\end{split}$$ {#eq-grpmean}
and similarly
$$\operatorname{clr} \bar{\pi}_{..}(x) = \sum_{\ell=1}^{d+m-1} \left(\frac{1}{G}
\sum_{g=1}^G \frac{1}{n_g} \sum_{i=1}^{n_g} \left[\pi_{gj}^{t}\right]^{C_\ell} \right) \cdot Z_\ell(x).$$ {#eq-globalmean}

As in @machalova_preprocessing_2016, after summarizing the initial
temperature samples using histograms, penalized least squares splines
are applied to smooth the data, treating the histogram bin centers as
the first coordinates and the corresponding relative frequencies as the
second coordinates. @fig-smoothing, which displays the province-level
temperature densities grouped by region, highlights distinct regional
patterns in the temperature distributions.

```{r fig-smoothing}
#| fig-cap: Densities of maximum temperature across the years 1987 to 2016.
#| fig-height: 5
vntry <- vietnam_temperature_dd |>
  left_join(vietnam_regions, by = c("region" = "code")) |>
  rename(region_name = name) |>
  arrange(region, province) |>
  select(!province) |>
  group_by(region, region_name, year) |>
  summarise(t_max = mean(t_max))

vntry |>
  plot_funs(t_max, color = year) +
  facet_wrap(vars(region), nrow = 2, axes = "all") +
  scale_linewidth(range = c(0.1, 1.2)) +
  scale_y_continuous(n.breaks = 4) +
	scale_color_viridis_c(end = 0.8) +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density",
    color = "Year"
  )
```

## Temporal trends by province {#sec-1sample_prov}

To capture the temporal evolution of temperature density distributions
across provinces within a given region (or within the whole Vietnam), we
consider a simplified trend model in which each province's density
evolves linearly over time in the Bayes space sense. More precisely, if
$\pi_{gj}^{t}$ now denotes the density in province $i$ ($i=1,\ldots, n$)
at time $t$, the simple time evolution model (for each province) is the
following density-on-scalar regression model:
$$\pi_{gj}^{t}(x)= [\alpha_{gj} \oplus ((t -t_0)\odot \beta_{gj}) \oplus \varepsilon_{gj}^{t}] (x),$$ {#eq-slopereg}
where $t_0 = 1987$ is the first observed date, $\beta_{gj}$ is the trend
slope density (or simply slope density) for province $i$ in region $g$.
The intercept $\alpha_{gj}$ is the theoretical density for province $j$
at time $t=t_0.$ When the slope density of this model is the uniform
distribution on $(a,b)$, there is no observable trend in the evolution
of temperature densities.

For estimating the parameters of @eq-slopereg, as in @talska_compositional_2018, the clr
transformation is applied to @eq-slopereg and the resulting equation is that of a simple
multivariate regression model which can be fitted by ordinary least
squares. The fitted slopes seem to give a good approximation of the
trend evolution since the coefficients of determination [defined in the
context of density-on-scalar regression in @talska_compositional_2018 p.
79] range from $63\%$ to $97\%$ (median of $94\%$) across provinces,
with a better fit in the south. As often done in functional data, we
will treat the resulting estimators of $\alpha_{gj}$ and $\beta_{gj}$ as
our sample density dataset (and therefore do not use the hat notation).
In @sec-intervalwise, we also use the trend density functions
${\tilde \pi}_{gj}$ of this model defined by
$${\tilde \pi}_{gj}^{t}(x) = [{\alpha}_{gj} \oplus (t\odot {\beta}_{gj})](x).$$ {#eq-mean_trend}
@fig-slopes displays the province specific slope densities ${\beta}_{gj}$ grouped by
region.

```{r fig-slopes}
#| fig-cap: Trends slopes of provinces grouped by region. The dotted line represents the uniform density on $[12^\circ\text C, 40^\circ\text C]$ which is the reference measure.
#| fig-height: 5
to_slope <- function(ddobj, t = seq_len(ncol(ddobj$coefs))) {
  model <- lm(t(ddobj$coefs) ~ t)

  # Density-on-scalar regression
  slope <- ddobj
  slope$coefs <- model$coefficients["t", ]

  # Compute functional R^2
  eps <- ddobj
  eps$coefs <- t(model$residuals)
  ssr_evol <- diag(inprod(eps, eps))
  sst_evol <- diag(inprod(ddobj, ddobj))
  ssr <- sum(ssr_evol)
  sst <- sum(sst_evol)

  slope <- dd(clr = fd(slope$coefs, slope$basis))
  attr(slope, "rsq") <- 1 - ssr / sst
  attr(slope, "ssr_evol") <- data.frame(t = t, ssr = ssr_evol / sst_evol)
  slope
}

vnt <- vietnam_temperature_dd |>
  group_by(region, province) |>
  summarise(t_max = list(c(t_max)), .groups = "drop") |>
  rowwise() |>
  mutate(t_max = list(to_slope(t_max))) |>
  mutate(rsq = attr(t_max, "rsq")) |>
  mutate(ssr_evol = list(attr(t_max, "ssr_evol"))) |>
  ungroup() |>
  left_join(vietnam_regions, by = c("region" = "code")) |>
  rename(region_name = name) |>
  arrange(region, province)
class(vnt$t_max) <- c("ddl", "fdl", "list")

vnt |> plot_funs(t_max, color = region_name) +
  facet_wrap(vars(region), nrow = 2, axes = "all") +
  scale_color_viridis_d(option = "magma", end = 0.8) +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  geom_hline(yintercept = 1 / diff(vnt$t_max[[1]]$basis$rangeval), linetype = "dotted") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )
```

In order to use our methodology, we will assume from now on that the
$\beta_{gj}$ are independent. Using the tools introduced by
@menafoglio_kriging_2014, we evaluated the spatial dependence between
the slope densities and found it negligible.

## Regional one-sample tests

Before testing the equality of trends between regions, it may be
interesting to perform a one-sample test in each region separately to
see whether there are some regions which are not affected by climate
change. For a given region $g$, we perform the one-sample test for the
trend slope sample of $\beta_{gj}$ for provinces $j$ in region $g$. To
evaluate the existence of climate change, we choose $\varphi_0$ to be
the neutral element (uniform density in the interval $[12,40]$), so that
in each region, the null hypothesis represents the absence of
temperature density change. The corresponding Šidák-adjusted $p$-values
[see @abdi_holms_2010] are displayed in  @tbl-one_sample_regional. This table strongly supports the
existence of climate change in the Red River Delta (RRD) region and in
the southern regions SR and MDR, the non-existence of climate change in
the Central Highlands (CHR) region. The results are more contrasted in
the Northern Midlands and Mountains (NMM) region and in the North
Central Coast (NCC) region where the tests are either non-significant or
very marginally significant. These results should be interpreted with
caution due to the small sample size in each region.

```{r one_sample_function}
one_sample <- function(ddobj, p = 3) {
  n <- ncol(ddobj$coefs)
  d <- nrow(ddobj$coefs)

  pca_fit <- pca.fd(ddobj, nharm = d)
  lambda <- pca_fit$values
  scores <- pca_fit$scores
  v <- pca_fit$harmonics

  # PC test
  tpc <- n * sum(inprod(v[1:p], mean(ddobj))^2 / lambda[1:p])
  tpc_pval <- pchisq(tpc, p, lower.tail = FALSE)

  # Norm test
  tnorm <- n * sum(inprod(v[lambda[1:d] > 0], mean(ddobj))^2)
  tnorm_pval <- imhof(tnorm, lambda[lambda[1:d] > 0])[[1]]

  one_sample_tbl <- tibble(
    Name = c("PC", "Norm"),
    Statistic = c(tpc, tnorm),
    `p-value` = c(tpc_pval, tnorm_pval)
  )
}
```

```{r tbl-one_sample_regional}
#| tbl-cap: Šidák-adjusted $p$-values for the regional one sample test.
one_sample_regional_tbl <- vnt |>
  select(region, region_name, t_max) |>
  summarise(t_max = list(c(t_max)), .by = c(region, region_name)) |>
  rowwise() |>
  mutate(res = list(one_sample(t_max))) |>
  unnest(res) |>
  select(region, Name, `p-value`) |>
  pivot_wider(names_from = Name, values_from = `p-value`)

multiple_testing <- function(tbl, m) {
  mutate(tbl, across(where(is.numeric), ~ 1 - (1 - .x)^m))
}

kable_format(multiple_testing(one_sample_regional_tbl,
  m = nrow(one_sample_regional_tbl)
))
```

## Regional distributional ANOVA {#sec-reg_dist_anova}

```{r fig-slopes_regional_means}
#| fig-cap: Regional mean trend slopes. The dotted line represents the uniform density on $[12^\circ\text C, 40^\circ\text C]$ which is the reference measure.
#| fig-height: 5
vntr <- vnt |>
  select(!province) |>
  group_by(region, region_name) |>
  summarise(t_max = mean(t_max))

vntr |> plot_funs(t_max, color = region_name) +
  geom_hline(yintercept = 1 / diff(vnt$t_max[[1]]$basis$rangeval), linetype = "dotted") +
  facet_wrap(vars(region), nrow = 2, axes = "all") +
  scale_color_viridis_d(option = "magma", end = 0.8) +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )
```

@fig-slopes_regional_means displays the regional mean trend
slope densities. For the central regions (CHR and NCC), this plot shows
a relative stability of the maximum temperature distribution (the trend
is not far from uniform). For the RRD region, the curve exhibits a
noticeable bump above the uniform on the right tail. More details about
the interpretation of these curves will be given in  @sec-intervalwise.

Using the tests of @sec-anova we now test whether these trend densities vary
across regions. It is indeed important to evaluate the null hypotheses
with different points of view due to the functional complex nature of
the densities: group means can differ in many ways. 
@tbl-slopes_danova_global summarizes
the results of the global analysis of variance for which the above tests
all conclude that there is a difference in the way the regional
temperature densities evolve in time.

```{r tbl-slopes_danova_global}
#| tbl-cap: Test statistics and $p$-values for DANOVA.
vnte <- vnt |>
  mutate(t_max = as.fd(t_max)) |>
  eval_funs(t_max, n = 1401) |>
  select(province, x, y) |>
  pivot_wider(names_from = province, values_from = y) |>
  column_to_rownames("x")

vnt_province_region <- vietnam_provinces$region
names(vnt_province_region) <- vietnam_provinces$province

stat_list <- list("GPF", "Fmaxb", "CS", "L2B", "FB")

res <- fanova.tests(vnte, vnt_province_region[colnames(vnte)],
  test = stat_list,
  params = list(
    paramCS = 100,
    paramFmaxb = 100,
    paramTRP = list(B.TRP = 100)
  )
)

danova_tbl <- res[as_vector(stat_list)] |>
  sapply(function(x) c(`Statistic` = x$stat, `p-value` = x$pval)) |>
  t() |>
  as.data.frame() |>
  rownames_to_column("Name")

kable_format(danova_tbl)
```

## Pairwise comparisons between regions

@tbl-slopes_danova_pairwise displays the tests statistics with
their Šidák-adjusted $p$-values for the five pairwise tests. We
summarize and visualize these results on @fig-slopes_danova_pairwise_diagram by drawing

1.  a solid line between two regions when none of the five tests reject
    the equality of mean slope densities
2.  a dotted line between two regions when the equality of mean slope
    densities is rejected by some but not all of the five tests
3.  no line between two regions when the equality of mean slope
    densities is rejected by all five tests.

At the exception of four dotted lines among the fifteen possible pairs,
the large number of solid lines and "no line" supports the fact that the
five tests agree most of the time. Note that the statistics GPF and
$F_{\max}$ agree in all cases but one. Among the statistics
$\operatorname{L^2B}$, FB and CS, there is at most one disagreement
between pairs.

The structure of @fig-slopes_danova_pairwise_diagram is interestingly similar
to the geographical structure of the regions (see the map on @fig-map) with solid
lines between neighboring regions. It shows that the Mekong Delta region
differs statistically from most other regions in terms of climate
change. Among the remaining regions, the Red River Delta region differs
the most from the others in the sense that it is only connected by
dotted lines.

```{r tbl-slopes_danova_pairwise}
#| tbl-cap: Šidák-adjusted $p$-values for pairwise group mean comparison.
vntp <- t(combn(c("NMM", "RRD", "NCC", "CHR", "SR", "MDR"), m = 2))
colnames(vntp) <- c("region1", "region2")

pairwise_anova <- function(region1, region2) {
  data <- vnte[vnt_province_region[colnames(vnte)] %in% c(region1, region2)]

  res <- fanova.tests(data, vnt_province_region[colnames(data)],
    test = stat_list,
    params = list(
      paramCS = 100,
      paramFmaxb = 100,
      paramTRP = list(B.TRP = 100)
    )
  )

  res[as_vector(stat_list)] |>
    sapply(function(x) c(x$pval)) |>
    c()
}

multiple_testing <- function(tbl, m) {
  mutate(tbl, across(where(is.numeric), ~ 1 - (1 - .x)^m))
}

pairwise_tbl <- as_tibble(vntp) |>
  rowwise() |>
  mutate(res = map2(region1, region2, pairwise_anova)) |>
  unnest_wider(res)

kable_format(multiple_testing(pairwise_tbl, m = nrow(pairwise_tbl)))
```

::: {.content-visible when-format="html"}
```{mermaid}
%%| label: fig-slopes_danova_pairwise_diagram
%%| fig-cap: "Summary of pairwise tests between regions of Vietnam. Solid line: no test rejects the null hypothesis; dotted line: some tests reject the null hypothesis; absence of line: all tests reject the null hypothesis."
%%| out-width: 60%
flowchart TB
  NMM([NMM]) ~~~ MDR([MDR])
  NMM <-.-> SR([SR])
  NMM <--> CHR([CHR]) <--> SR
  NMM <-.-> RRD([RRD]) <-.-> CHR
  RRD <--> NCC
  NCC <--> CHR
  NMM <--> NCC([NCC]) <--> SR
  SR <-.-> MDR
  NMM ~~~ MDR([MDR])

style NMM fill:#000004,color:#fff,stroke:#000
style RRD fill:#29115b,color:#fff,stroke:#000
style NCC fill:#6d1c82,color:#fff,stroke:#000
style CHR fill:#ac337b,stroke:#000
style SR fill:#ea5662,stroke:#000
style MDR fill:#ff9f6f,stroke:#000
```
:::

::: {.content-visible unless-format="html"}
![Summary of pairwise tests between regions of Vietnam. Solid line: no test rejects the null hypothesis; dotted line: some tests reject the null hypothesis; absence of line: all tests reject the null hypothesis.](fig-slopes_danova_pairwise_diagram.pdf){#fig-slopes_danova_pairwise_diagram height=45%}
:::

## Interpreting regional changes in temperature distributions {#sec-regional_interpretation}

Averaging @eq-mean_trend over the province index for a given region
$g$, we get the regional trend density 
$${\tilde \pi}_{g.}^{t}(x) = [{\alpha}_{g.} \oplus (t\odot {\bar \beta}_{g.})](x),$$ {#eq-mean_trend2}
and therefore the negative perturbation between two consecutive (between
time $t$ and $t+1$) trend density functions is constant equal to the
regional slope density $\bar\beta_{g.}.$ Hence as in @sec-intervalwise, we can relate the relative frequency of
$\beta_{g.}$ at two temperature points $x$ and $z$ to the infinitesimal
odds ratio of ${\tilde \pi}_{g.}^{t+1}$ versus ${\tilde \pi}_{g.}^{t}$

```{r interval_functions}
fmin <- function(...) UseMethod("fmin")
fmax <- function(...) UseMethod("fmax")
fargmin <- function(...) UseMethod("fargmin")
fargmax <- function(...) UseMethod("fargmax")

fmin.fdl <- function(...) fmin(c(...))
fmax.fdl <- function(...) fmax(c(...))
fargmin.fdl <- function(...) fargmin(c(...))
fargmax.fdl <- function(...) fargmax(c(...))

fsolve <- function(...) UseMethod("fsolve")
fsolve.fdl <- function(...) fargmax(c(...))

fmin.fd <- function(fdobj,
                    a = min(fdobj$basis$rangeval),
                    b = max(fdobj$basis$rangeval)) {
  f <- function(x) eval(fdobj, x)
  results <- lapply(
    seq_along(f(a)),
    function(i) optimize(function(x) f(x)[i], interval = c(a, b))
  )
  sapply(results, `[[`, "objective")
}

fargmin.fd <- function(fdobj,
                       a = min(fdobj$basis$rangeval),
                       b = max(fdobj$basis$rangeval)) {
  f <- function(x) eval(fdobj, x)
  results <- lapply(
    seq_along(f(a)),
    function(i) optimize(function(x) f(x)[i], interval = c(a, b))
  )
  sapply(results, `[[`, "minimum")
}

fmax.fd <- function(fdobj,
                    a = min(fdobj$basis$rangeval),
                    b = max(fdobj$basis$rangeval)) {
  f <- function(x) -eval(fdobj, x)
  results <- lapply(
    seq_along(f(a)),
    function(i) optimize(function(x) f(x)[i], interval = c(a, b))
  )
  -sapply(results, `[[`, "objective")
}

fargmax.fd <- function(fdobj,
                       a = min(fdobj$basis$rangeval),
                       b = max(fdobj$basis$rangeval)) {
  f <- function(x) -eval(fdobj, x)
  results <- lapply(
    seq_along(f(a)),
    function(i) optimize(function(x) f(x)[i], interval = c(a, b))
  )
  sapply(results, `[[`, "minimum")
}

fsolve.fd <- function(fdobj, k,
                      a = min(fdobj$basis$rangeval),
                      b = max(fdobj$basis$rangeval)) {
  f <- function(x) eval(fdobj, x) - k
  results <- lapply(
    seq_along(f(a)),
    function(i) uniroot(function(x) f(x)[i], interval = c(a, b))
  )
  sapply(results, `[[`, "root")
}
```

```{r fig-interval_mountains}
#| fig-cap: Mean slope densities of North Central Coast (NCC) and Northern Midlands and Mountains (NMM) regions with levels.
#| fig-height: 3
thres1 <- 0.5 * (fmax(vntr$t_max[[1]], 30, 35) + fmin(vntr$t_max[[1]], 20, 30))
thres3 <- 0.5 * (fmax(vntr$t_max[[3]], 30, 35) + fmin(vntr$t_max[[3]], 20, 30))

vnt_mountains_thresholds <- rbind(tibble(
  region = "NMM",
  x = c(
    fsolve(vntr$t_max[[1]], fmax(vntr$t_max[[1]], 30, 35), 20, 25),
    fsolve(vntr$t_max[[1]], thres1, 20, 25),
    fsolve(vntr$t_max[[1]], thres1, 25, 30),
    fsolve(vntr$t_max[[1]], thres1, 30, 35),
    fsolve(vntr$t_max[[1]], fmin(vntr$t_max[[1]], 20, 30), 30, 35)
  ),
  y = c(
    fmax(vntr$t_max[[1]], 30, 35),
    thres1,
    thres1,
    thres1,
    fmin(vntr$t_max[[1]], 20, 30)
  )
), tibble(
  region = "NCC",
  x = c(
    fsolve(vntr$t_max[[3]], fmax(vntr$t_max[[3]], 30, 35), 15, 25),
    fsolve(vntr$t_max[[3]], thres3, 20, 25),
    fsolve(vntr$t_max[[3]], thres3, 25, 30),
    fsolve(vntr$t_max[[3]], thres3, 30, 35),
    fsolve(vntr$t_max[[3]], fmin(vntr$t_max[[3]], 25, 30), 30, 35)
  ),
  y = c(
    fmax(vntr$t_max[[3]], 30, 35),
    thres3,
    thres3,
    thres3,
    fmin(vntr$t_max[[3]], 25, 30)
  )
))

vntr |>
  filter(region %in% c("NMM", "NCC")) |>
  plot_funs(t_max) +
  geom_vline(data = vnt_mountains_thresholds, aes(xintercept = x, color = y, linetype = -y)) +
  geom_hline(data = vnt_mountains_thresholds, aes(yintercept = y, color = y, linetype = -y)) +
  facet_wrap(vars(region), nrow = 1, axes = "all") +
  scale_linetype_binned() +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )
```

```{r tbl-interval_mountains}
#| tbl-cap: Levels for North Central Coast (NCC) and Northern Midlands and Mountains (NMM) regions.
vnt_mountains_thresholds |> mutate(x = signif(x, 3), y = signif(y, 3))
```

```{r fig-interval_rrd}
#| fig-cap: Mean slope densities of Red River Delta (RRD) and Central Highlands (CHR) regions with levels.
#| fig-height: 3
vnt_rrd_thresholds <- rbind(tibble(
  region = "RRD",
  x = c(
    fsolve(vntr$t_max[[2]], fmax(vntr$t_max[[2]], 12, 15), 35, 37),
    fsolve(vntr$t_max[[2]], fmax(vntr$t_max[[2]], 12, 15), 37, 40),
    fsolve(vntr$t_max[[2]], fmin(vntr$t_max[[2]], 37, 40), 15, 20),
    fsolve(vntr$t_max[[2]], fmin(vntr$t_max[[2]], 37, 40), 30, 35)
  ),
  y = c(
    fmax(vntr$t_max[[2]], 12, 15),
    fmax(vntr$t_max[[2]], 12, 15),
    fmin(vntr$t_max[[2]], 37, 40),
    fmin(vntr$t_max[[2]], 37, 40)
  )
), tibble(
  region = "CHR",
  x = c(
    fsolve(vntr$t_max[[4]], fmax(vntr$t_max[[4]], 20, 30), 35, 40),
    fsolve(vntr$t_max[[4]], fmin(vntr$t_max[[4]], 30, 40), 15, 20),
    fsolve(vntr$t_max[[4]], fmin(vntr$t_max[[4]], 30, 40), 20, 25)
  ),
  y = c(
    fmax(vntr$t_max[[4]], 20, 30),
    fmin(vntr$t_max[[4]], 30, 40),
    fmin(vntr$t_max[[4]], 30, 40)
  )
))

vntr |>
  filter(region %in% c("RRD", "CHR")) |>
  plot_funs(t_max) +
  geom_vline(data = vnt_rrd_thresholds, aes(xintercept = x, color = y, linetype = -y)) +
  geom_hline(data = vnt_rrd_thresholds, aes(yintercept = y, color = y, linetype = -y)) +
  facet_wrap(vars(region), nrow = 1, axes = "all") +
  scale_linetype_binned() +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )
```

```{r tbl-interval_rrd}
#| tbl-cap: Levels for Red River Delta (RRD) and Central Highlands (CHR) regions.
vnt_rrd_thresholds |> mutate(x = signif(x, 3), y = signif(y, 3))
```

```{r fig-interval_south}
#| fig-cap: Mean slope densities of Southeast (SR) and Mekong Delta (MDR) regions with levels.
#| fig-height: 3
vnt_south_thresholds <- rbind(tibble(
  region = "SR",
  x = c(
    fsolve(vntr$t_max[[5]], fmax(vntr$t_max[[5]], 12, 20), 20, 25),
    fsolve(vntr$t_max[[5]], fmax(vntr$t_max[[5]], 12, 20), 25, 30),
    fsolve(vntr$t_max[[5]], fmin(vntr$t_max[[5]], 12, 25), 30, 35)
  ),
  y = c(
    fmax(vntr$t_max[[5]], 12, 20),
    fmax(vntr$t_max[[5]], 12, 20),
    fmin(vntr$t_max[[5]], 12, 25)
  )
), tibble(
  region = "MDR",
  x = c(
    fsolve(vntr$t_max[[6]], fmax(vntr$t_max[[6]], 12, 20), 25, 30),
    fsolve(vntr$t_max[[6]], fmax(vntr$t_max[[6]], 12, 20), 30, 35),
    fsolve(vntr$t_max[[6]], fmin(vntr$t_max[[6]], 12, 25), 30, 35)
  ),
  y = c(
    fmax(vntr$t_max[[6]], 12, 20),
    fmax(vntr$t_max[[6]], 12, 20),
    fmin(vntr$t_max[[6]], 12, 25)
  )
))

vntr |>
  filter(region %in% c("SR", "MDR")) |>
  plot_funs(t_max) +
  geom_vline(data = vnt_south_thresholds, aes(xintercept = x, color = y, linetype = -y)) +
  geom_hline(data = vnt_south_thresholds, aes(yintercept = y, color = y, linetype = -y)) +
  facet_wrap(vars(region), nrow = 1, axes = "all") +
  scale_linetype_binned() +
  scale_y_continuous(n.breaks = 4) +
  theme(legend.position = "none") +
  labs(
    x = "Temperature (deg. Celsius)",
    y = "Density"
  )
```

```{r tbl-interval_south}
#| tbl-cap: Levels for Southeast (SR) and Mekong Delta (MDR) regions.
vnt_south_thresholds |> mutate(x = signif(x, 3), y = signif(y, 3))
```

@fig-interval_mountains, @fig-interval_rrd and @fig-interval_south
display the slope densities of the six
regions with some chosen values of level $\tau$. We are able to group
the regions in terms of the shape of their mean slope density.

On @fig-interval_mountains for northern regions, we select some
levels $\tau$ of the regional slope density $\beta_{g.}$ and picture
them on the graph: they are chosen so as to divide the full range
(alternatively part of it) into two subintervals where the slope density
is above (respectively below) that level or the reverse. For the NCC
region for example, the highest level separates the temperature range
into temperatures below 20$^\circ\text C$ and temperatures above
20$^\circ\text C$. The corresponding curve $\beta_{g.}$ is above the
chosen level for low temperatures and below for temperatures higher than
20$^\circ\text C$, therefore by @eq-atau we
may conclude that there is an increase of the frequency of low
temperatures over time. Note that all levels higher than the previous
one lead to the same conclusion. Similarly, the lowest level of $\tau$
reveals a decrease of values of temperature approximately above
$33.7^\circ\text C$. Focusing now solely on the medium range between
21$^\circ\text C$ and 33$^\circ\text C$, the intermediate level
separates the medium range into the interval
$[21^\circ\text C,29.1^\circ\text C]$ where the slope density is below
the level and its complement (within the medium range) where the slope
density is above the level. We can conclude that in the medium range
there is a shift towards higher temperatures over time. The same
conclusion holds for the NMM region for which the medium range is
$[22.6^\circ\text C,27.6^\circ\text C]$.

With the same strategy, based on @fig-interval_rrd, we can see that the RRD region displays
an increasing spread of its temperature distribution over time (values
of $\tau$ below the lowest level) and an increase of extremely high
temperature events (values of $\tau$ above the highest level). Note that
since temperature change is not detected in the CHR region (@tbl-slopes_danova_global), its
interpretation based on odds ratio is probably not reliable.

Based on @fig-interval_south, the south in contrast shows a
concentration in the medium range around 25$^\circ\text C$ for SR and
28$^\circ\text C$ for MDR at the expense of high temperatures.

While these groups reflect the geography (latitude and elevation) of
Vietnam, they are also consistent with the groups suggested by the
regional one-sample tests (@tbl-one_sample_regional) and the pairwise comparisons
(@tbl-slopes_danova_pairwise).

# Conclusion

We have adapted several functional data analysis tests to density
functions in order to assess the equality of mean densities and to
perform DANOVA tests in the framework of Bayes spaces. In our target
application of temperature density evolution in Vietnam, the one-sample
test allows to conclude that there is statistical evidence of the
existence of climate change in Vietnam in the sense that the trend slope
density is not uniform. Furthermore, when considering the functions
globally, most of the DANOVA statistics strongly reject the null
hypothesis, that is, the equality of trend slope density across the
Vietnamese regions. For more local interpretations, we rely on the
infinitesimal odds ratio of @maier_additive_2025.

Needless to say that the proposed methodology can be applied in other
application frameworks. In our application, the inference has been
carried out without taking into account the fact that the slope
distributions are estimated with a linear model from data rather than
observed. An alternative solution for testing the uniformity of climate
change across regions would be to use a linear model for explaining the
densities by time and a regional effect and to test the parameters of
this model. In order to take into account the possible spatial
dependence of the data set studied in this article, we could consider
adapting the Spatial FANOVA methodology introduced by
@aristizabal_analysis_2019 to a distributional context.

Future research might try to reduce the uncertainty due to the low
density of extreme temperatures, possibly weighting the domain of the
Bayes space as in @talska_weighting_2020. Our local investigations
remain exploratory, and elevating them to formal tests presents an
interesting avenue for future research.

Our methodology is promising for environmental studies, with a possible
application for example to the relative concentrations of contaminants
in grounds or rivers analyzed as density functions.

# Acknowledgments {.appendix .unnumbered}

Part of this work was completed while the authors were visiting the
Vietnam Institute for Advanced Study in Mathematics (VIASM) in Hanoi and
the authors express their gratitude to VIASM. We also acknowledge
funding from the French National Research Agency (ANR) under grant
ANR-17-EURE-0010 (Investissements d'Avenir program), from the French
National Association of Research and Technology (ANRT), grant CIFRE
n°2020/0011, and from Spanish Ministerio de Ciencia e Innovación grant
number PID2021-123833OB-I00.
